---
title: "estimateRtintervention"
author: "Hanyang Liu"
date: "6/9/2020"
output: html_document
---

```{r}
library(rstan)
library(ggplot2)
library(dplyr)
library("bayesplot")
theme_set(bayesplot::theme_default(base_family = "sans"))
library(loo)
library(gridExtra)
```


1. Obtain the infection potential 

```{r serial interval}
inf_pot <- function(locnum, impnum, day){
  '
  function to find infection potential
  
  arguments: 
  locnum/ impnum: local and imported incidents
                  (day + 1) rownum and 1 colnum dataframe   `
                  
                  
  day: one dimentional vector (0-indexed)
         
  return: 
  one dimention infection potential on day t
  '
  
  totnum = locnum + impnum
  omega_mat = c()
  for (n in 1:day){
    # use recommended serial distribution
    omega = pgamma(n+0.5, shape = 6.5, rate = 0.62) - pgamma(n-0.5, shape = 6.5, rate = 0.62)
    omega_mat[n] = omega
  }
  # normalize the data
  omega_mat = omega_mat/ (1-pgamma(0.5, shape = 6.5, rate = 0.62))
  omega_mat = matrix(omega_mat)
  totnum_mat = matrix(totnum[day:1,1], nrow = 1,  ncol = day)
  infectionpot = c(totnum_mat %*% omega_mat)
  return(infectionpot)
}
```



2. Perform Bayesian Analysis

```{r Bayesian Analysis}
covidRt <- function(locnum, impnum, day, start_shape, start_rate, timewindow, CI){
  '
  function to find Rt, l and u according to CI on certain day
  
  arguments: 
  locnum/ impnum: (day + 1) rownum and 1 colnum dataframe
  day: one dimention (0-indexed)
  start_shape/start_rate: parameters for prior
  timewindow: reproduction number is constant over [t-timewindow+1, t]
  CI: probability of credible interval
  
  return:
  MEAN Rt value on day, Lambda, lower bound of CI, upper bound of CI
  newshape, newrate
  '
  
  locnum_sum = 0
  infpot_sum = 0
  for (n in (day-timewindow+1):day){
    locnum_sum = locnum_sum + c(locnum[n+1,1])
    infpot_sum = infpot_sum + inf_pot(locnum, impnum, n)
  }
  newshape = start_shape + locnum_sum
  newrate = start_rate + infpot_sum
  # we take mean of poisson distribution as Rt
  Rt = newshape / newrate
  # l and u for CI
  Lambda = inf_pot(locnum, impnum, day)
  l = qgamma((1-CI)/2, shape = newshape, rate = newrate)
  u = qgamma(1-(1-CI)/2, shape = newshape, rate = newrate)
  # update shape and rate
  return(c(Rt, Lambda, l, u, newshape, newrate))
}

```

3. Obtaining summary statistics

```{r summary}
covidRt_summary <- function(df, startshape = 1, 
                            startrate = 0.2, tau = 1, ci = 0.95){
  '
  function to add Rt, l and u columns to dataset
  
  arguments:
  df: 2(3) columns dataframe: Date, local_daily_increase, (import_daily_increase) in order
  startshape: prior shape
  startrate: prior rate
  tau: timewindow to make reproduction number constant over [t-tau+1, t]
  ci: credible interval
  
  return:
  updated dataframe with new columns Rt, lower and upper
  '
  
  # to ensure column numbers as required
  stopifnot(ncol(df)>=2, ncol(df)<=3)
  # add import incidents if missing
  if (ncol(df) == 2){
    df$import_daily = 0
  }
  # initiate values
  Rt_list = rep(0,tau)
  l_list = rep(0,tau)
  u_list = rep(0,tau)
  Lambda_list = rep(0,tau)
  shape_list = rep(0,tau)
  rate_list = rep(0,tau)
  daynum = nrow(df) - 1
  # for loop to update values
  for (iterations in tau:daynum){
    parameter = covidRt(locnum = df[2], 
                        impnum = df[3], 
                        day = iterations,
                        timewindow = tau,
                        start_shape  = startshape,
                        start_rate = startrate,
                        CI = ci)
    Rt_list[iterations+1] = parameter[1]
    Lambda_list[iterations+1] = parameter[2]
    l_list[iterations+1] = parameter[3]
    u_list[iterations+1] = parameter[4]
    shape_list[iterations+1] = parameter[5]
    rate_list[iterations+1] = parameter[6]
  }
  # add columns to dataframe
  df$Rt = Rt_list
  df$Lambda = Lambda_list
  df$lower = l_list
  df$upper = u_list
  df$expect_incidence = df$Rt * df$Lambda
  df$shape = shape_list
  df$rate = rate_list
  return(df)
}
```

simple preprocessing function

```{r preprocessing function}
cumutodaily <- function(df){
  '
  Arguments: 
  df: single column dataframe representing cumulative infection number
  
  Return:
  preprocessed dataframe
  '
  # to clean levels in dataframe
  df = as.numeric(as.character(df))
  # subtraction
  for (n in length(df):2){
    df[n] = df[n] - df[n-1]
  }
  return(df)
}

preprocessing <- function(df_date, df_confirmed){
  '
  arguments:
  df: dataframe
  df_date: dataframe reprerents column of date
  df_confirmed: dateframe represents column of cumulative confirmed cases
  
  return:
  preprocessed dataframe
  '
  pproc_cumu_confirmed = cumutodaily(df_confirmed)
  pproc_df = data.frame(date = df_date, incidents = pproc_cumu_confirmed)
  return(pproc_df)
}
```


```{r preprocessing global data}
# read csv from your file location
df = read.csv('time_series_covid19_confirmed_global.csv', check.names=FALSE)
region = paste(df$'Country/Region', df$'Province/State', sep = '')
df = df[-(1:4)]
df = data.frame(t(df))
df = add_rownames(df, var = "date")
df$date <- as.Date(df$date)
colnames(df)[2:ncol(df)] = region
df
```

new function starts from now:

a preprocessing function for intervention model

```{r}
transform_intervention_df <- function(date, region){
  df = covidRt_summary(preprocessing(date, region),
                       tau = 1,
                       )
  df = data.frame(date = df$date, incidents = df$incidents, Lambda = df$Lambda)
  df = df[-1,]
  rownames(df) <- NULL
  return(df)
}
```

functions used to initiate data for stan functions

```{r}
intervention_days_select <- function(datedf, intervention_days){
  '
  '
  validdays = c()
  for (day in 1:length(intervention_days)){
    if (any(datedf == intervention_days[day])){
      validdays = c(validdays, c(intervention_days[day]))
    } else{
      print(sprintf('%s is not in our date range', intervention_days[day]))
    }
  }
  return(validdays)
}

initiate_dat = function(intervention_df, intervention_days, alphavar = 1, betavar = 1,
                        alpha0mean = 0, alpha0var = 1){
  incidents = as.integer(intervention_df$incidents)
  infection_potential = intervention_df$Lambda
  N = length(incidents)
  intervention_days = intervention_days_select(intervention_df$date, intervention_days)
  intervention_times = length(intervention_days)
  stopifnot(intervention_times > 0)
  intervention = matrix(0, nrow = N, ncol = intervention_times)
  for (n in 1:intervention_times){
    intervention[match(intervention_days[n], as.character(intervention_df$date)):N,n] = 1
  }
  alphasd = sqrt(alphavar)
  betasd = sqrt(betavar)
  alpha0sd = sqrt(alpha0var)

  return(list(N = N,
              incidents = incidents,
              infec_pot = infection_potential,
              intervention_times = intervention_times,
              intervention = intervention, 
              alphasd = alphasd,
              betasd = betasd,
              alpha0mean = alpha0mean,
              alpha0sd = alpha0sd))
}
```

I have checked Chinese websites, there are some wrong data. have corrected below

```{r}
df$ChinaHubei[22] = 48206
df$ChinaHubei[23] = 51986
df$ChinaHubei[32] = 63454
df$ChinaHubei
```

```{r}
hubei_df = transform_intervention_df(df$date, df$ChinaHubei)
hubei_df[86,2] = 0   # 4-17 anormaly
```

2020-02-12 looks weird, how are we gonna deal with them?


we need to change our model parameter below to produce a better result

```{r}
hubei_intervention_days = c('2020-01-27', '2020-03-25', '2020-04-08')
hubei_dat = initiate_dat(intervention_df = hubei_df[11:nrow(hubei_df),], # start from 2020-02-02 as over 4000?
                         intervention_days = hubei_intervention_days,
                         # definition for parameters below are same as in stan file
                         alphavar = 0.01,      
                         betavar = 0.01,
                         alpha0mean = 0.50515, # log(3.2) 3.2 is in CI of R0 in Hubei in many papers
                         alpha0var = 1)
```


```{r}
stan_model1 = 'Rtintervention.stan'
hubei_fit <- stan(file = stan_model1,
            data = hubei_dat, 
            iter = 10000, 
            chains = 4,
            # control=list(adapt_delta=0.8),
            )
```



plot for Rt_mean with 95% CI as ribbon. 

```{r}
Rt_mean = c()
Rt_sd = c()
Rt_lower = c()
Rt_upper = c()
for (n in 1:length(hubei_dat$incidents)){
  Rt_mat = as.matrix(hubei_fit, pars = paste('Rt[', as.character(n),']'))
  Rt_mean[n] = mean(Rt_mat)
  Rt_sd[n] = sd(Rt_mat)
  Rt_lower[n] = quantile(Rt_mat, probs = 0.025)
  Rt_upper[n] = quantile(Rt_mat, probs = 0.975)
}
hubei_Rt_df = data.frame(day = (1:length(hubei_dat$incidents))-1,
                         Rt_mean = Rt_mean, 
                         Rt_sd = Rt_sd, 
                         Rt_approx = hubei_dat$incidents/hubei_dat$infec_pot,
                         incidents = hubei_dat$incidents)
hubei_Rt_plot = ggplot(data = hubei_Rt_df, aes(x = day, y = Rt_mean))+geom_path()
hubei_Rt_plot = hubei_Rt_plot + geom_ribbon(aes(ymin = Rt_lower, ymax = Rt_upper))
hubei_Rt_plot = hubei_Rt_plot + geom_hline(aes(yintercept = 1))
hubei_Rt_plot = hubei_Rt_plot + ylim(0,7)
hubei_Rt_plot
```
```{r}
print(hubei_fit, pars = "beta")
```
```{r}
hubei_Rt_plot1 = ggplot(data = hubei_Rt_df, aes(x = day, y = incidents))+geom_path()
hubei_Rt_plot1
grid.arrange(hubei_Rt_plot, hubei_Rt_plot1, ncol=2)
```

test our model quality by generating incidence_rep based on poisson(Rt*Lambdat)

```{r}
incidents_rep = as.matrix(hubei_fit, pars = 'incidents_rep')
dim(incidents_rep)
```

change t which is day t

```{r}
t = 25
I_rep_df = data.frame(incidents_rep)
hubei_Rt_plot2 = ggplot(data = I_rep_df, aes(x = I_rep_df[,t+1])) 
hubei_Rt_plot2 = hubei_Rt_plot2 + geom_histogram(binwidth=5, fill = 'lightblue') 
hubei_Rt_plot2 = hubei_Rt_plot2 + geom_vline(aes(xintercept = hubei_dat$incidents[t+1]), color = 'red')
hubei_Rt_plot2 = hubei_Rt_plot2 + labs(x = 'incidents',
                                       y = 'count',
                                       title = 'Model Quality Check')
hubei_Rt_plot2
```



anormaly

```{r}
df$`United Kingdom`[120] = (df$'United Kingdom'[119]+df$'United Kingdom'[121])/2
df$`United Kingdom`
```

Now we can deal with UK 

```{r}
UK_df = transform_intervention_df(df$date, df$`United Kingdom`)

```



we need to change our model parameter below to produce a better result

```{r}
UK_intervention_days = c('2020-03-23','2020-05-10','2020-06-01')
UK_dat = initiate_dat(intervention_df = UK_df[58:nrow(UK_df),], # start from 2020-03-20 as over 4000?
                         intervention_days = UK_intervention_days,
                         # definition for parameters below are same as in stan file
                         alphavar = 1,      
                         betavar = 1,
                         alpha0mean = 0.3617, # log(2.3) 2.3 is in CI of R0 in UK in March?
                         alpha0var = 1)
```

also try change iter?

```{r}
stan_model1 = 'Rtintervention.stan'
UK_fit <- stan(file = stan_model1,
            data = UK_dat, 
            iter = 30000, 
            chains = 4,
            control = list(max_treedepth = 15)
            # control=list(adapt_delta=0.8),
            )
```

plot for Rt_mean with 95% CI as ribbon]

```{r}
Rt_mean = c()
Rt_sd = c()
Rt_lower = c()
Rt_upper = c()
for (n in 1:length(UK_dat$incidents)){
  Rt_mat = as.matrix(UK_fit, pars = paste('Rt[', as.character(n),']'))
  Rt_mean[n] = mean(Rt_mat)
  Rt_sd[n] = sd(Rt_mat)
  Rt_lower[n] = quantile(Rt_mat, probs = 0.025)
  Rt_upper[n] = quantile(Rt_mat, probs = 0.975)
}
UK_Rt_df = data.frame(day = (1:length(UK_dat$incidents))-1,
                         Rt_mean = Rt_mean, 
                         Rt_sd = Rt_sd, 
                         Rt_approx = UK_dat$incidents/UK_dat$infec_pot)
UK_Rt_plot = ggplot(data = UK_Rt_df, aes(x = day, y = Rt_mean))+geom_path()
UK_Rt_plot = UK_Rt_plot + geom_ribbon(aes(ymin = Rt_lower, ymax = Rt_upper))
UK_Rt_plot = UK_Rt_plot + geom_hline(aes(yintercept = 1))
UK_Rt_plot = UK_Rt_plot + ylim(0,13)
UK_Rt_plot
```


```{r}
print(UK_fit, pars = 'beta', probs = c(0.05, 0.95))

```
test our model quality by generating incidence_rep based on poisson(Rt*Lambdat)

```{r}
incidents_rep = as.matrix(UK_fit, pars = 'incidents_rep')
dim(incidents_rep)
```


```{r}
t = 25
I_rep_df = data.frame(incidents_rep)
UK_Rt_plot2 = ggplot(data = I_rep_df, aes(x = I_rep_df[,t+1])) 
UK_Rt_plot2 = UK_Rt_plot2 + geom_histogram(binwidth=5, fill = 'lightblue') 
UK_Rt_plot2 = UK_Rt_plot2 + geom_vline(aes(xintercept = UK_dat$incidents[t+1]), color = 'red')
UK_Rt_plot2 = UK_Rt_plot2 + labs(x = 'incidents',
                                       y = 'count',
                                       title = 'Model Quality Check')
UK_Rt_plot2
```





Now we can deal with Greece

```{r}
Greece_df = transform_intervention_df(df$date, df$Greece)

```



we need to change our model parameter below to produce a better result

```{r}
Greece_intervention_days = c('2020-02-27', '2020-03-10', '2020-03-13','2020-03-22','2020-05-04','2020-05-11')
Greece_dat = initiate_dat(intervention_df = Greece_df[43:nrow(Greece_df),], # start from 2020-03-05?
                         intervention_days = Greece_intervention_days,
                         # definition for parameters below are same as in stan file
                         alphavar = 1,      
                         betavar = 1,
                         alpha0mean = 0.3617, # log(2.3) 2.3 simply guessing
                         alpha0var = 1)
```

also try change iter?

```{r}
stan_model1 = 'Rtintervention.stan'
Greece_fit <- stan(file = stan_model1,
            data = Greece_dat, 
            iter = 10000, 
            chains = 4,
            control = list(max_treedepth = 15)
            # control=list(adapt_delta=0.8),
            )
```

plot for Rt_mean with s.d. as ribbon.

```{r}
Rt_mean = c()
Rt_sd = c()
for (n in 1:length(Greece_dat$incidents)){
  Rt_mat = as.matrix(Greece_fit, pars = paste('Rt[', as.character(n),']'))
  Rt_mean[n] = mean(Rt_mat)
  Rt_sd[n] = sd(Rt_mat)
}
Greece_Rt_df = data.frame(day = (1:length(Greece_dat$incidents))-1,
                         Rt_mean = Rt_mean, 
                         Rt_sd = Rt_sd, 
                         Rt_approx = Greece_dat$incidents/Greece_dat$infec_pot)
Greece_Rt_plot = ggplot(data = Greece_Rt_df, aes(x = day, y = Rt_mean))+geom_path()
Greece_Rt_plot = Greece_Rt_plot + geom_ribbon(aes(ymin = Rt_mean-Rt_sd, ymax = Rt_mean+Rt_sd))
Greece_Rt_plot = Greece_Rt_plot + geom_hline(aes(yintercept = 1))
Greece_Rt_plot = Greece_Rt_plot + ylim(0,5)
Greece_Rt_plot
```
```{r}
print(Greece_fit, pars = 'Rt')
```



test our model quality by generating incidence_rep based on poisson(Rt*Lambdat)

```{r}
incidents_rep = as.matrix(Greece_fit, pars = 'incidents_rep')
dim(incidents_rep)
```

```{r}
t = 25
I_rep_df = data.frame(incidents_rep)
Greece_Rt_plot2 = ggplot(data = I_rep_df, aes(x = I_rep_df[,t+1])) 
Greece_Rt_plot2 = Greece_Rt_plot2 + geom_histogram(binwidth=2, fill = 'lightblue') 
Greece_Rt_plot2 = Greece_Rt_plot2 + geom_vline(aes(xintercept = Greece_dat$incidents[t+1]), color = 'red')
Greece_Rt_plot2 = Greece_Rt_plot2 + labs(x = 'incidents',
                                       y = 'count',
                                       title = 'Model Quality Check')
Greece_Rt_plot2
```











